### 2021-07-02 ch/2-related-work.tex
Fix typos/language problems/grammar errors

The first works in the field almost exclusively took the definition-centric view,
most of them using \acl{DP} \citep{dwork2012fairness} as the fairness criterion.

The boundary between these views is not completely clear-cut.
Consider the case where a biased training set is available,
and an unbiased test set is assumed to exist, but we do not have access to it \citep{jiang2020identifying}.
Then we might at least know that the unbiased test set satisfies \acf{DP}.
In this case, testing the classifier for \ac{DP} is an indication
for how it would fare on the unbiased test set.

@@

\section{Foundations of algorithmic fairness}\label{sec:foundations}
This section discusses the publications that lay the ground work for the algorithmic fairness literature.
More recent publications are discussed in \secref{sec:refinement-on-previous-ideas-for-fair-classification-and-fair-representations}.

\subsection{Precursors}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data-mining.
Their first fundamental conclusion
is that it is not sufficient to remove the protected attribute
(which should not be used as a basis for prediction) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features%
\footnote{This is similar to the legal view of direct and indirect discrimination
where direct discrimination when people with a specific protected characteristic are treated worse,
whereas indirect discrimination is a policy that does not make direct references to a protected characteristic
but disproportionally affects people with a specific characteristic.}.
% The latter case can be
% detected by checking the other rules in the dataset.

@@

The authors formulate the problem as it was later done in other works on fairness as well:
there is a set of features \(x\), a binary class label \(y\)
and a special \emph{sensitive attribute} \(s\) which determines the demographic group.
The sensitive attribute \(s\) is assumed to be binary in the derivation,
but does not have to be.
\(s=0\) refers to the group vulnerable to discrimination and \(s=1\) to all other individuals.
It is further assumed that one of the class labels is generally desirable;
for example, it might correspond to being accepted for a loan or being given bail.
This class label is being referred to as the positive label: \(y = 1\).

@@

Their algorithm for de-biasing the dataset comprises two steps.
First, a well-calibrated classifier (Na\"ive Bayes in the paper) is trained on the data
\emph{without} the sensitive attribute.
This classifier is then used to determine a ranking for how likely a positive label is for a given individual.
The highest ranked individuals that have label \(y=0\)
and are potentially discriminated against \(s=0\) get ``promoted'' to \(y=1\)
and the lowest ranked individuals with label \(y=1\) and \(s=1\) get ``demoted'' to \(y=0\)
until the dataset is fair according to the \(Disc\) measure.
The idea being that the ranking ensures that the changes in the dataset
happen to the most appropriate candidates (those closest to the decision boundary);
the main goal being a prevention of a big drop in accuracy of the final classifier
on the test set whose distribution matches the training set.

@@

The second method in \citet{calders2010three}
is based on training two separate models for each sensitive attributes.
That is, the dataset is split in two
where one half contains all examples with \(s=0\), we call this dataset \(D_0\),
and the other half contains all examples with \(s=1\), called \(D_1\).
For prediction, we choose either the model that was trained on \(D_0\) or the one for \(D_1\),
depending on the sensitive attribute of the example.
This method relies on the availability of the sensitive attribute for prediction,
like the previous method.
The two models are separately tweaked to produce overall fair results.
In order to minimise the effect on the accuracy, the models are tweaked the same amount each,
but in opposite directions.
Conceptually, having two different Na\"ive Bayes model depending on the (binary) sensitive attribute
is equivalent to using one Na\"ive Bayes model
in which the sensitive attribute is connected to all other features.
This means that the bias is in all of the features and not just in the class label.
The latter was the assumption of the previously discussed model.
The difference between these assumptions is not explored in much detail in the paper.

The third method introduces a new latent variable, called \(L\) in the paper.
\(L\) is intended to be an unbiased target class label
that replaces the biased class label \(y\) from the training data.
\(L\) can only be estimated and is unbiased
in the sense that it is statistically independent from sensitive attribute: \(P(L|s=0) = P(L|s=1)\).
The observed class label \(y\) then only depends on the latent label and the sensitive attribute.
\(L\) is found by using expectation maximisation
to iteratively search for a value that maximises the likelihood of the dataset.
The search is restricted by enforcing that \(L\) must be equal to \(y\) except
when \(s=0\) and \(y=0\) or when \(s=1\) and \(y=1\)
because these are the two cases where we expect discrimination.
There is other prior knowledge that can be incorporated into the search.
% This method seems to have been the most sophisticated at that point in time.

\citet{calders2010three} present experimental results for the three methods on an artificial dataset
and the Adult/Census Income data that \citet{calders2009building} used before.
The artificial data has biased and also unbiased class labels
which are generally not available for real-world datasets
and conforms to the assumptions made for the third model.
% Artificial datasets generally have the advantage that the bias can be controlled very precisely,
% but there is always the danger of tailoring the dataset too closely to the proposed algorithm.
% To the authors surprise,
On both datasets, the two simpler methods outperform the third method
that is based on fair latent class labels;
the two first methods perform approximately equally well.
% The authors mention the possibility of a better fairness measure.

@@

\begin{enumerate}
\item
  \emph{Prejudice}:
  a statistical dependency between the sensitive attribute
  and the class label or the (non-sensitive) input features.
  If the dependency is with the class label, then the paper calls it \emph{indirect prejudice}.
  If it is with the non-sensitive features, they call it \emph{latent prejudice}.
  In line with \citet{pedreshi2008discrimination},
  \emph{direct prejudice} refers to classifiers that make direct use of the sensitive attribute.
  Indirect prejudice and latent prejudice can be quantified by
  the mutual information between \(y\) and \(s\) or \(x\) and \(s\) respectively.
  \citet{kamishima2012fairness} do not consider latent prejudice to be immediately harmful,
  but it can be a problem with respect to the protection of personal data and compliance with laws.
\item
  \emph{Underestimation}:
  non-convergence of machine learning algorithm due to limited amount of data.
  This magnitude of this problem can be estimated
  by considering the difference between the actual training sample distribution
  and the distribution that the machine learning model has internalised.
  This is the underlying cause of models that make predictions
  that are even more unfair than the dataset.
\item
  \emph{Negative Legacy}:
  sampling bias and wrong labels in the dataset.
  In contrast to the problem of \emph{prejudice},
  \emph{negative legacy} might not be detectable by analysing the dataset.
  \emph{Sampling bias}, meaning that certain data points simply are missing,
  and \emph{wrong labels} can only be corrected if other sources of information are available.
\end{enumerate}

We see that previous works nearly exclusively dealt with the first of these issues, \emph{prejudice},
and more precisely \emph{indirect prejudice}.
\citet{kamishima2012fairness} also focus on \emph{indirect prejudice};
essentially because it is the easiest to deal with, apart from \emph{direct prejudice}.
They use \acf{LR} as the basis for their fair classifier.
Their approach to enforcing fairness is qualitatively different from the previous proposals.
Instead of manipulating the dataset or using explicit algorithm to make a classifier fair,
\citet{kamishima2012fairness} treat the fairness constraint like a regulariser.
A term is added to the objective function
that estimates the mutual information between \(y\) and \(s\)
and gets minimised together with the other parts of the objective function.
% This strategy leaves most of the work to the optimisation that would need to be run anyway.
A factor in front of the regularisation term determines how much to value fairness over accuracy.

@@

\subsection{Fairness based on similarity}%
\label{fairness-based-on-similarity}
Nearly all previously discussed papers express some dissatisfaction
with the \acf{DP} fairness definition.
\citet{luong2011k} use a very different fairness metric
based on the idea that similar people should be treated similarly.
The setup is still that the training set is biased,
only a different fairness definition is used to judge classifiers.
The authors first define a distance metric which defines a neighbourhood,
based on the Manhattan distance on the z-scores of the attributes.
The distance metric is supposed to be only making use of legally admissible attributes.
An individual is then considered to be unfairly treated
if it is classified differently than its neighbours.
More concretely, for any data point we can check
how many of the \(k\) nearest neighbours have the same class label as that data point.
If the percentage is under a certain threshold
then there was discrimination against the individual corresponding to that data point.
In order to create a fair classifier, for this definition of fairness,
the authors propose to pre-process the dataset,
similar to \citet{kamiran2009classifying} before,
by flipping the class labels of those data points where the class label is considered unfair.
The experiments in the paper show that this method is successful in reducing the discrimination
(according to the given criterion)
for a range of classifiers on the Adult/Census Income dataset.

\citet{dwork2012fairness} give a more in-depth analysis
of the idea of distance-based fairness and \ac{DP}.
They present explicit detailed criticism of the \ac{DP} metric,
mainly in the form of several scenarios in which \ac{DP} is maintained,
but the system treats a lot of individuals very unfairly.
In one scenario, the sensitive attribute carries important information
and removing it makes everyone worse off.
% In the second scenario, a malicious actor achieves \ac{DP}
% by promoting the weakest members of the disadvantaged group to receive a positive label
% and then argues that fairness is not a worthwhile goal
% because these individuals will not be able to succeed.
% The goal of the actor here is to undermine the whole goal of avoiding discrimination.
Another scenario considers the case of \emph{subset targeting},
where there is discrimination against subgroups
which are not covered by the \ac{DP} metric
that only considers the major demographic groups.
Defining fairness via group membership will always leave open the possibility
of discriminating against ever smaller subgroups down to the level  of individuals.

@@

\subsection{Fair representations}\label{sec:fair-representation}
\citet{zemel2013learning} approach the problem of biased training data more directly
by seeking to transform the features such that all traces of the sensitive attribute are removed,
but so that they still carry the information necessary for predicting the class label.
% \citet{zemel2013learning} expand on the idea of creating a \emph{fair representation}.
% That is, the features of the data points are transformed such that all bias is removed
This is different from \citet{kamiran2009classifying} where the dataset was transformed as well,
but there, the change was to the \emph{labels}.
The idea of transforming the features assumes that the main problem with the data
are unwanted dependencies between $s$ and the other features (called \emph{latent prejudice} in \citet{kamishima2012fairness}),
and that the training labels are mostly unproblematic.
The intended result is then that classifiers trained on the transformed data will make $s$-invariant predictions `by default',
because they do not know about $s$.

More specifically, the stated goal of \citet{zemel2013learning} is
to achieve both \emph{group fairness} (in the sense of Demographic Parity)
and \emph{individual fairness} (in the sense of treating similar people similarly),
as \citet{dwork2012fairness} did before.
Let \(z\) denote the learned fair representation.
The condition for fairness is then
\begin{align}
  \label{eq:fair-representation}
  P(s=0|z=z') = P(s=1|z=z') ~.
\end{align}
\citet{zemel2013learning} propose to map the biased inputs \(x\) to \emph{prototypes} in the same space as \(x\).
The probability for being assigned to a particular prototype
must be the same for inputs with \(s=0\) and for \(s=1\).
Based on a given distance function (or similarity measure),
inputs are more likely to be assigned to prototypes that are close-by.
The classification task is then based entirely on the prototypes (the fair representation \(z\)).
In their work, a linear model is used to map the prototypes to the outcomes \(y\).
Aside from the distance function, the whole model has therefore only two kinds of parameters:
the locations of the prototypes and the weights in the linear model.
These parameters are all optimised together.
The objective function consists of three terms:
the first enforcing Demographic Parity via the prototype locations,
the second ensuring that the prototypes are close to the inputs
and the third targeting a good accuracy by adapting the weights of the classifier.
%
The experiments are done on the German credit dataset, the Adult / Census Income dataset
and a dataset based on the Heritage Health Prize milestone 1 challenge~\citep{heritagehealth}
where the goal is to predict how many days a given person will spend in the hospital in a year.
% The proposed methods performs better in these experiments
% than the methods from \citet{kamiran2009classifying} and the method from \citet{kamishima2012fairness}.
% However, they do not compare against \citet{calders2010three},
% presumably because those methods rely on the sensitive attribute for making predictions,
% which can be considered an unfair advantage.

@@

\citet{feldman2015certifying} try to improve upon the work by \citet{zemel2013learning},
focusing on fair representation as well.
The authors try to ground their definition of fairness in U.S.\ law.
They define \emph{Disparate Impact} (DI) as
\begin{align}
  \label{eq:disparate-impace}
  DI = \frac{P(y=1|s=0)}{P(y=1|s=1)} ~.
\end{align}
With a target of \(DI = 1\), this would simply enforce \ac{DP}.
However, based on some rulings and recommendations of the U.S.\ legal system,
they advocate the 80\% Rule, which states that \(DI\) should not be below 80\% (or above 125\%).
This means that the acceptance rate of a disadvantaged demographic group
should not be less than 80\% of the acceptance rate of the other group.
With this definition, there is an explicit allowed range for small unfairness.
Previously, people just tried to get as close as possible to \(DI = 1\),
without stating how close is close enough.

For measuring the fairness of the (non-sensitive) input features
(disregarding class labels for the moment), the authors define \(\varepsilon\)-fairness.
The features \(x\) are \(\varepsilon\)-fair,
if any predictor that tries to predict \(s\) from \(x\) can only achieve a balanced error rate
that is higher than \(\varepsilon\).
\citet{feldman2015certifying} prove that that \(\varepsilon\)-fairness with a suitable \(\varepsilon\)
is incompatible with violating the 80\% rule.
That is, if the sensitive attribute cannot be predicted from the features,
then a classifier trained on that data will automatically be fair (to a certain degree).
For datasets that are extremely unbalanced,
the required \(\varepsilon\) approaches 1/2
which corresponds to absolutely no information in \(x\) about \(s\).
Note that the definitions require that the performance of the best possible classifiers is known,
which is rarely the case.
In the paper, an \ac{SVM} classifier is used to measure the \(\varepsilon\) for \(\varepsilon\)-fairness.

In order to create a fair dataset,
the authors present an algorithm that considers every feature individually
and shifts the values such that the distributions \(P(z=z'|s=0)\) and \(P(z=z'|s=1)\) are identical
(\(z\) refers to the shifted values, \(x\) refers to the original values).
This shift retains the ordering of the data points with regard to that feature.
This is to ensure that \(z\) can still be used to predict \(y\).
The method only works on numerical features.
The paper contains two other algorithms for removing unfairness which are not as invasive,
meaning some amount of unfairness remains, but the ability to predict is improved.
They are referred to as \emph{Partial Repair} algorithms, as opposed to full repair.
This is the fairness-accuracy trade-off that basically all works in this area consider.
Both of the other methods try to preserve the ranking of the data points
with respect to the individual features
while minimising the distance of the distributions for the different groups.

@@

A plausible question is whether it is possible achieve all of these criteria simultaneously.
The authors show that a perfect predictor,
which gives a score of \(P(\hat{y}=1|x) = 1\) to data points with \(y=1\)
and \(P(\hat{y}=0|x) = 1\) to those with \(y=0\),
automatically satisfies all three criteria.
(The same is in general not true for \ac{DP}:
if the test set labels exhibit a statistical dependency to $s$, then a perfect predictor does the same.)
Furthermore, the authors show that if the dataset is fair in the sense that it satisfies \(P(y=1|s=0) = P(y=1|s=1)\),
\ie, the base acceptance rate is the same for both classes,
then a ``random'' classifier which assigns that base rate as the prediction score to all data points indiscriminately
also satisfies all three criteria.
For this case, \ac{DP} is satisfied as well.
The paper provides a proof
that those two cases are the only ones that achieve the three presented guarantees simultaneously.
Demographic Parity can only be achieved simultaneously with these in the second scenario.
The general case sits between those extremes:
the predictor is not perfect and the test set is not unbiased,
and therefore, these definitions of fairness are not compatible.
Note, however, that criterion 2 and 3 \emph{are} compatible with one another.

@@

\citet{hardt2016equality} expand on criterion 3 in \citet{kleinberg2016inherent}
and develop a method to enforce it via post-processing.
Furthermore, an alternative criterion is introduced that is the combination of criterion 2 and 3.
This is also the work that gave criterion 3 the name \acl{EOpp}
and that popularised the name \acl{DP}.
The treatment of fairness by the authors is explicitly ``oblivious'',
which here means that only the general statistics are known
about the features \(x\), the sensitive attributes and the class labels \(y\).
In particular, there is not enough information available to develop a similarity measure
which could be used to target \emph{individual fairness}.

\citet{hardt2016equality} define the criteria in terms of statistical independence:
\acf{EOdds} refers to the case where \(\hat{y}\) and \(s\) are independent conditional on \(y\).
This is equivalent to the true positive rate and the false positive rate being the same for all groups.
\ac{EOpp} is, as mentioned above,
the case where just the true positive rate is the same for all groups.
This means that \(\hat{y}\) and \(s\) are independent conditional on \(y=1\).
One main advantage that these definitions have over \ac{DP} is that a perfect predictor satisfies them on any evaluation set.
% Demographic Parity is at odds with achieving perfect accuracy.
% Furthermore, for Demographic Parity a classifier might be forced to give a positive prediction to a lot of individuals
% where it is really not appropriate.

>>>>

The first works in the field almost exclusively took the definition-centric view,
most of them using \acf{DP} \citep{dwork2012fairness} as the fairness criterion.

The boundary between these views is not completely clear-cut.
Consider the case where a biased training set is available,
and an unbiased test set is assumed to exist, but we do not have access to it \citep{jiang2020identifying}.
Then we might at least know that the unbiased test set satisfies \ac{DP}.
In this case, testing the classifier for \ac{DP} is an indication
for how well it would fare on the unbiased test set.

@@

\section{Foundations of algorithmic fairness}\label{sec:foundations}
This section discusses the publications that lay the ground work for the algorithmic fairness literature.
More recent publications are discussed in %
\secref{sec:refinement-on-previous-ideas-for-fair-classification-and-fair-representations}.

\subsection{Precursors}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data-mining.
Their first fundamental conclusion
is that it is not sufficient to remove the protected attribute
(which should not be used as a basis for prediction) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features%
\footnote{This is similar to the legal view of direct and indirect discrimination
where direct discrimination is when people with a specific protected characteristic are treated worse,
whereas indirect discrimination is a policy that does not make direct references to a protected characteristic
but disproportionally affects people with a specific characteristic.}.
% The latter case can be
% detected by checking the other rules in the dataset.

@@

The authors formulate the problem as it was later done in other works on fairness:
there is a set of features \(x\), a binary class label \(y\)
and a special \emph{sensitive attribute} \(s\) which determines the demographic group.
The sensitive attribute \(s\) is assumed to be binary in the derivation,
but does not have to be.
\(s=0\) refers to the group vulnerable to discrimination and \(s=1\) to all other individuals.
It is further assumed that one of the class labels is generally desirable;
for example, it might correspond to being accepted for a loan or being given bail.
This class label is referred to as the positive label: \(y = 1\).

@@

Their algorithm for de-biasing the dataset comprises two steps.
First, a well-calibrated classifier (Na\"ive Bayes in the paper) is trained on the data
\emph{without} the sensitive attribute.
This classifier is then used to determine a ranking for how likely a positive label is for a given individual.
The highest ranked individuals that have label \(y=0\)
and are potentially discriminated against (\(s=0\)) get ``promoted'' to \(y=1\)
and the lowest ranked individuals with label \(y=1\) and \(s=1\) get ``demoted'' to \(y=0\)
until the dataset is fair according to the \(Disc\) measure.
The idea being that the ranking ensures that the changes in the dataset
happen to the most appropriate candidates (those closest to the decision boundary);
the main goal being a prevention of a big drop in accuracy of the final classifier
on the test set whose distribution matches the training set.

@@

The second method in \citet{calders2010three}
is based on training two separate models for each sensitive attribute.
That is, the dataset is split in two
where one half contains all examples with \(s=0\), we call this dataset \(D_0\),
and the other half contains all examples with \(s=1\), called \(D_1\).
For prediction, we choose either the model that was trained on \(D_0\) or the one for \(D_1\),
depending on the sensitive attribute of the example.
This method relies on the availability of the sensitive attribute for prediction,
like the previous method.
The two models are separately tweaked to produce overall fair results.
In order to minimise the effect on the accuracy, the models are tweaked the same amount each,
but in opposite directions.
Conceptually, having two different Na\"ive Bayes model depending on the (binary) sensitive attribute
is equivalent to using one Na\"ive Bayes model
in which the sensitive attribute is connected to all other features.
This means that the bias is in all of the features and not just in the class label.
The latter was the assumption of the previously discussed model.
The difference between these assumptions is not explored in much detail in the paper.

The third method introduces a new latent variable, called \(L\) in the paper.
\(L\) is intended to be an unbiased target class label
that replaces the biased class label \(y\) from the training data.
\(L\) can only be estimated and is unbiased
in the sense that it is statistically independent from the sensitive attribute: \(P(L|s=0) = P(L|s=1)\).
The observed class label \(y\) then only depends on the latent label and the sensitive attribute.
\(L\) is found by using expectation maximisation
to iteratively search for a value that maximises the likelihood of the dataset.
The search is restricted by enforcing that \(L\) must be equal to \(y\) except
when \(s=0\) and \(y=0\) or when \(s=1\) and \(y=1\)
because these are the two cases where we expect discrimination.
There is other prior knowledge that can be incorporated into the search.
% This method seems to have been the most sophisticated at that point in time.

The authors present experimental results for the three methods on an artificial dataset
and the Adult/Census Income data that \citet{calders2009building} used before.
The artificial data has biased and also unbiased class labels,
the latter of which are generally not available for real-world datasets,
and the data conforms to the assumptions made for the third model.
(The existence of the unbiased ground-truth labels makes this paper an example of the ground-truth-centric view.)
% Artificial datasets generally have the advantage that the bias can be controlled very precisely,
% but there is always the danger of tailoring the dataset too closely to the proposed algorithm.
% To the authors surprise,
On both datasets, the two simpler methods outperform the third method
that is based on fair latent class labels;
the two first methods perform approximately equally well.
% The authors mention the possibility of a better fairness measure.

@@

\begin{enumerate}
\item
  \emph{Prejudice}:
  a statistical dependency between the sensitive attribute
  and the class label or the (non-sensitive) input features.
  If the dependency is with the class label, then the paper calls it \emph{indirect prejudice}.
  If it is with the non-sensitive features, they call it \emph{latent prejudice}.
  In line with \citet{pedreshi2008discrimination},
  \emph{direct prejudice} refers to classifiers that make direct use of the sensitive attribute.
  \citet{kamishima2012fairness} do not consider latent prejudice to be immediately harmful,
  but it can be a problem with respect to the protection of personal data and compliance with laws.
\item
  \emph{Underestimation}:
  non-convergence of machine learning algorithm due to limited amount of data.
  This magnitude of this problem can be estimated
  by considering the difference between the actual training sample distribution
  and the distribution that the machine learning model has internalised.
  This is the underlying cause of models that make predictions
  that are even more unfair than the dataset.
\item
  \emph{Negative Legacy}:
  sampling bias and wrong labels in the dataset.
  In contrast to the problem of \emph{prejudice},
  \emph{negative legacy} might not be detectable by analysing the dataset.
  \emph{Sampling bias}, meaning that certain data points simply are missing,
  and \emph{wrong labels} can only be corrected if other sources of information are available.
\end{enumerate}

We see that previous works nearly exclusively dealt with the first of these issues, \emph{prejudice},
and more precisely \emph{indirect prejudice}.
\citet{kamishima2012fairness} also focus on \emph{indirect prejudice};
mostly because it is the easiest to deal with, apart from \emph{direct prejudice}.
They use \acf{LR} as the basis for their fair classifier.
Their approach to enforcing fairness is qualitatively different from the previous proposals.
Instead of manipulating the dataset or using explicit algorithm to make a classifier fair,
\citet{kamishima2012fairness} treat the fairness constraint like a regulariser.
A term is added to the objective function
that estimates the mutual information between \(y\) and \(s\)
and gets minimised together with the other parts of the objective function.
% This strategy leaves most of the work to the optimisation that would need to be run anyway.
A factor in front of the regularisation term determines how much to value fairness over accuracy.

@@

\subsection{Fairness based on similarity}%
\label{fairness-based-on-similarity}
Nearly all previously discussed papers express some dissatisfaction
with the \acf{DP} fairness definition.
\citet{luong2011k} use a very different fairness metric
based on the idea that similar people should be treated similarly.
The setup is still that the training set is biased,
only a different fairness definition is used to judge classifiers.
The authors first define a distance metric which defines a neighbourhood,
based on the Manhattan distance on the z-scores of the attributes.
The distance metric is supposed to be only making use of legally admissible attributes.
An individual is then considered to be unfairly treated
if it is classified differently than its neighbours.
More concretely, for any data point we can check
how many of the \(k\) nearest neighbours have the same class label as that data point.
If the percentage is under a certain threshold
then -- so the authors argue -- there was discrimination against the individual corresponding to that data point.
In order to create a fair classifier, for this definition of fairness,
the authors propose to pre-process the dataset,
similar to \citet{kamiran2009classifying} before,
by flipping the class labels of those data points where the class label is considered wrong or biased.
The experiments in the paper show that this method is successful in reducing the discrimination
(according to the given criterion)
for a range of classifiers on the Adult/Census Income dataset.

\citet{dwork2012fairness} give a more in-depth analysis
of the idea of distance-based fairness and \ac{DP}.
They present explicit detailed criticism of the \ac{DP} metric,
mainly in the form of several scenarios in which \ac{DP} is maintained,
but the system treats a lot of individuals very unfairly.
In one scenario, the sensitive attribute carries important information
and removing it makes everyone worse off.
% In the second scenario, a malicious actor achieves \ac{DP}
% by promoting the weakest members of the disadvantaged group to receive a positive label
% and then argues that fairness is not a worthwhile goal
% because these individuals will not be able to succeed.
% The goal of the actor here is to undermine the whole goal of avoiding discrimination.
Another scenario considers the case of \emph{subset targeting},
where there is discrimination against subgroups
which are not covered by the \ac{DP} metric
that only considers the major demographic groups.
(This problem has also been referred to as \emph{hidden stratification}.)
Defining fairness via group membership will always leave open the possibility
of discriminating against ever smaller subgroups down to the level  of individuals.

@@

\subsection{Fair representations}\label{sec:fair-representation}
\citet{zemel2013learning} approach the problem of biased training data more directly
by seeking to transform the features, such that all traces of the sensitive attribute are removed,
while at the same time letting the features still carry the information required for predicting the class label.
% \citet{zemel2013learning} expand on the idea of creating a \emph{fair representation}.
% That is, the features of the data points are transformed such that all bias is removed
This is different from \citet{kamiran2009classifying} where the dataset was transformed as well,
but there, the change was to the \emph{labels}.
The idea of transforming the features assumes that the main problem with the data
are unwanted dependencies between $s$ and the other features (called \emph{latent prejudice} in \citet{kamishima2012fairness}),
and that the training labels are mostly unproblematic.
The intended result is then that classifiers trained on the transformed data will make $s$-invariant predictions `by default',
because they do not know about $s$.

More specifically, the stated goal of \citet{zemel2013learning} is
to achieve both \emph{group fairness} (in the sense of Demographic Parity)
and \emph{individual fairness} (in the sense of treating similar people similarly),
as \citet{dwork2012fairness} did before.
Let \(z\) denote the learned fair representation.
The condition for fairness is then
\begin{align}
  \label{eq:fair-representation}
  P(s=s'|z=z') = P(s=s') \quad \forall s' \in \{0, 1\}, z' \in \mathbb{Z}~.
\end{align}
\citet{zemel2013learning} propose to map the biased inputs \(x\) to \emph{prototypes} in the same space as \(x\).
The probability for being assigned to a particular prototype
must be the same for inputs with \(s=0\) and for \(s=1\).
Based on a given distance function (or similarity measure),
inputs are more likely to be assigned to prototypes that are close-by.
The classification task is then based entirely on the prototypes (the fair representation \(z\)).
In their work, a linear model is used to map the prototypes to the outcomes \(y\).
Aside from the distance function, the whole model has therefore only two kinds of parameters:
the locations of the prototypes and the weights in the linear model.
These parameters are all optimised together.
The objective function consists of three terms:
the first enforcing \acf{DP} via the prototype locations,
the second ensuring that the prototypes are close to the inputs
and the third trying to maximise accuracy by adapting the weights of the classifier.
%
The experiments are done on the German credit dataset, the Adult / Census Income dataset
and a dataset based on the Heritage Health Prize milestone 1 challenge~\citep{heritagehealth}
where the goal is to predict how many days a given person will spend in the hospital in a year.
% The proposed methods performs better in these experiments
% than the methods from \citet{kamiran2009classifying} and the method from \citet{kamishima2012fairness}.
% However, they do not compare against \citet{calders2010three},
% presumably because those methods rely on the sensitive attribute for making predictions,
% which can be considered an unfair advantage.

@@

\citet{feldman2015certifying} try to improve upon the work by \citet{zemel2013learning},
focusing on fair representation as well.
The authors try to ground their definition of fairness in U.S.\ law.
They define \emph{Disparate Impact} (DI) as
\begin{align}
  \label{eq:disparate-impace}
  DI = \frac{P(y=1|s=0)}{P(y=1|s=1)} ~.
\end{align}
With a target of \(DI = 1\), this would simply enforce \ac{DP}.
However, based on some rulings and recommendations of the U.S.\ legal system,
they advocate the 80\% Rule, which states that \(DI\) should not be below 80\% (or above 125\%).
This means that the acceptance rate of a disadvantaged demographic group
should not be less than 80\% of the acceptance rate of the other group.
With this definition, there is an explicit allowed range for small unfairness.
Previously, researchers just tried to get as close as possible to \(DI = 1\),
without stating how close is close enough.

For measuring the fairness of the (non-sensitive) input features
(disregarding class labels for the moment), the authors define \(\varepsilon\)-fairness.
The features \(x\) are \(\varepsilon\)-fair,
if any predictor that tries to predict \(s\) from \(x\) can only achieve a balanced error rate
that is higher than \(\varepsilon\).
\citet{feldman2015certifying} prove that \(\varepsilon\)-fairness with a suitable \(\varepsilon\)
is incompatible with violating the 80\% rule.
That is, if the sensitive attribute cannot be predicted from the features,
then a classifier trained on that data will automatically be fair (to a certain degree).
For datasets that are extremely unbalanced,
the required \(\varepsilon\) approaches 1/2
which corresponds to absolutely no information in \(x\) about \(s\).
Note that the definitions require that the performance of the best possible classifiers is known,
which is rarely the case.
In the paper, \iac{SVM} classifier is used to measure the \(\varepsilon\) for \(\varepsilon\)-fairness.

In order to create a fair dataset,
the authors present an algorithm that considers every feature individually
and shifts the values such that the distributions \(P(z=z'|s=0)\) and \(P(z=z'|s=1)\) are identical
(\(z\) refers to the shifted values, \(x\) refers to the original values).
This shift retains the ordering of the data points with regard to that feature.
This is to ensure that \(z\) can still be used to predict \(y\)
(which assumes that \(y\) and \(s\) are sufficiently uncorrelated,
so that \(z\) can at the same time be unpredictive of \(s\) and predictive of \(y\)).
The method only works on numerical features.
The paper contains two other algorithms for removing unfairness which are not as invasive,
meaning some amount of unfairness remains, but the ability to predict is improved.
They are referred to as \emph{Partial Repair} algorithms, as opposed to full repair.
This is the fairness-accuracy trade-off that basically all works in this area consider.
Both of the other methods try to preserve the ranking of the data points
with respect to the individual features
while minimising the distance of the distributions for the different groups.

@@

A plausible question is whether it is possible to achieve all of these criteria simultaneously.
The authors show that a perfect predictor,
\ie one that gives a score of \(P(\hat{y}=1|x) = 1\) to data points with \(y=1\)
and \(P(\hat{y}=0|x) = 1\) to those with \(y=0\),
automatically satisfies all three criteria.
(The same is in general not true for \ac{DP}:
if the test set labels exhibit a statistical dependency to $s$, then a perfect predictor does the same.)
Furthermore, the authors show that if the dataset is fair in the sense that it satisfies \(P(y=1|s=0) = P(y=1|s=1)\),
\ie, the base acceptance rate is the same for both classes,
then a ``random'' classifier which assigns that base rate as the prediction score to all data points indiscriminately
also satisfies all three criteria.
For this case, \ac{DP} is satisfied as well.
The paper provides a proof
that those two cases are the only ones that achieve the three presented guarantees simultaneously.
Demographic Parity can only be achieved simultaneously with these in the second scenario (``random classifier'').
The general case sits between those extremes:
the predictor is not perfect and the test set is not unbiased,
and therefore, these definitions of fairness are not compatible.
Note, however, that criterion 2 and 3 \emph{are} compatible with one another.

@@

\citet{hardt2016equality} expand on criterion 3 in \citet{kleinberg2016inherent}
and develop a method to enforce it via post-processing.
Furthermore, an alternative criterion is introduced that is the combination of criterion 2 and 3.
This is also the work that gave criterion 3 the name \emph{\acl{EOpp}}
and that popularised the name \emph{\acl{DP}}.
The treatment of fairness by the authors is explicitly ``oblivious'',
which here means that only the general statistics are known
about the features \(x\), the sensitive attributes and the class labels \(y\).
In particular, there is not enough information available to develop a similarity measure
which could be used to target \emph{individual fairness}.

The authors define the criteria in terms of statistical independence:
\acf{EOdds} refers to the case where \(\hat{y}\) and \(s\) are independent conditional on \(y\).
This is equivalent to the true positive rate and the false positive rate being the same for all groups.
\ac{EOpp} is, as mentioned above,
the case where just the true positive rate is the same for all groups.
This means that \(\hat{y}\) and \(s\) are independent conditional on \(y=1\).
One main advantage that these definitions have over \ac{DP} is that a perfect predictor satisfies them on any evaluation set.
% Demographic Parity is at odds with achieving perfect accuracy.
% Furthermore, for Demographic Parity a classifier might be forced to give a positive prediction to a lot of individuals
% where it is really not appropriate.
